---
title: "Santa Clara County Educational Attainment Indicators"
output: flexdashboard::flex_dashboard
---

```{r global, include=F}
knitr::opts_chunk$set(echo = TRUE)
library(flexdashboard)
library(tidyverse)
library(leaflet)
library(plotly)
library(devtools)
devtools::install_github("jamgreen/lehdr")
library(readxl)
library(tigris)
library(sf)
library(lehdr)
Sys.setenv(CENSUS_KEY="c8aa67e4086b4b5ce3a8717f59faa9a28f611dab")

scc_per_change <- readRDS("scc_per_change.rds")
scc_pums_example <- readRDS("scc_pums_example.rds")
scc_pums <- readRDS("scc_pums.rds")
scc_pumas <- readRDS("scc_pumas.rds")
```

Column
-------------------------------------

The map below shows discrepancies of up to 9% in between adjacent geographic 
areas in SCC. Interestingly, the center of San Jose has a lower percentage of 
students without internet access--this analysis can be extended more broadly 
to identify if it is a pattern. These discrepancies point to the difficulties 
of transitioning to remote learning for a small percentage of K-12 students. 
These discrepancies may indicate compounding inequities for this subset of 
students (secure housing, financial stability, presence of caregivers at home). 
This map is an initial dive into persisting educational disparities. To help 
inform funding priorities regarding internet access, a more specific and 
comprehensive analysis would be of use.

### % K-12 Student Internet Access

```{r, echo=F}
scc_pums_example <-
  scc_pums %>% 
  filter(!duplicated(SERIALNO)) %>% 
  mutate(
    internet_access = ifelse(
      (ACCESS %in% c("1","2")),
      PWGTP,
      0
    )
  ) %>% 
  group_by(PUMA) %>% 
  summarize(
    perc_no_internet =
      (100 - sum(internet_access, na.rm =T)/sum(PWGTP, na.rm = T)*100)
  ) %>% 
  left_join(
    scc_pumas %>% 
      select(PUMACE10),
    by = c("PUMA" = "PUMACE10")
  ) %>% 
  st_as_sf()

pums_pal <- colorNumeric(
  palette = "plasma",
  domain = scc_pums_example$perc_no_internet
)

leaflet() %>%
  addTiles() %>% 
  addPolygons(
    data = scc_pums_example,
    fillColor = ~pums_pal(perc_no_internet),
    color = "white",
    opacity = 0.5,
    fillOpacity = 0.5,
    weight = 1,
    label = ~paste0(
      round(perc_no_internet), 
      "% K-12 students without internet access"
    ),
    highlightOptions = highlightOptions(
      weight = 2,
      opacity = 1
    )
  ) %>% 
  addLegend(
    data = scc_pums_example,
    pal = pums_pal,
    values = ~perc_no_internet,
    title = "% K-12 students<br>without internet access<br>in Santa Clara County"
  )
```

Column
-------------------------------------

This graph shows the percent change across educational outcomes, from high 
school graduates to 4-year college graduates, broken down by race for Santa 
Clara county (SCC). The Asian population in SCC has 
the greatest percent change out of any racial group; yet, are many 
disparities between different Asian communities based on ethnicity (e.g. 
Southeast Asian and East Asian communities), that this graph’s data set (ACS) 
lacks. The negative percent changes for the Indigenous populations in SCC is 
concerning. There is a clear discrepancy in education attainment between 
Indigenous groups and other racial groups, at least in terms of higher 
education, and the causes are complex. Info on how this dashboard 
was made: http://github.com/natram3/natram3.github.io/blob/master/R-9.rmd


### % Change From High School Graduates to Bachelor’s Degrees


```{r, echo=F}
scc_per_change %>% 
  ggplot() +
  geom_bar(
    aes(
      x = race %>% factor(),
      y = change2,
      fill = race
    ),
    stat = "identity",
    position = "stack"
  ) +
  labs(
    x = "Race",
    y = "% Change",
    title = "% Change from High School Graduates to Bachelor's",
    fill = "Race"
  ) + 
  coord_flip() + 
  theme(legend.position = "bottom", legend.direction = "vertical")
```
